---
title: "HW3_tidy_data_instructions"
author: "Aidan O'Hara"
date: "2023-09-20"
output: pdf_document
header-includes:
  - \usepackage{float}
  - \usepackage{subfig}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
```

### Instructions

Tidy the following tables from the Hadley Wickham, Tidy Data article into the given tidy (and melt) forms.

Only use the libraries included in the starter code: 

```{r eval = F}
library(foreign) #read.spss
library(stringr)
library(dplyr)
library(tidyr)
```

Use the included `dim(df) == c(x,y)` tests to check you have the right size data frame. 

### PREG

The starter code will the generate the raw data needed for the `preg` set. Use this as a good reminder that small portions of a given dataset can be used to first develop the tidy form before attempting to tidy the dataset at large.

```{=tex}
\begin{table}[H]
  \centering
  \input{data/preg-raw-1.tex}
  \caption{Typical presentation dataset.}
  \label{tbl:preg-raw-1}
\end{table}
```
```{=tex}
\begin{table}[H]
  \centering
  \input{data/preg-tidy.tex}
  \caption{The same data as in Table~\ref{tbl:preg-raw-1} but with variables in columns and observations in rows.}
  \label{tbl:preg-tidy}
\end{table}
```
\newpage

### PEW

This next table gets much more involved.

```{=tex}
\begin{table}[H]
  \centering
  \input{data/pew-raw.tex}
  \caption{The first ten rows of data on income and religion from the Pew Forum. Three columns, \texttt{\$75-100k}, \texttt{\$100-150k} and \texttt{>150k}, have been omitted} 
  \label{tbl:pew-raw}
\end{table}
```

You will need to coerce some strings into the specified forms present in the tidy table. 

Specifically, you will need to trim the `reltrad` column of `pew` from:

`[1]  Evangelical Protestant Churches`\
`[2]  Mainline Protestant Churches`\
`[3]  Unaffiliated`\
`[4]  Jewish`\
`[5]  Don’t know/refused (no information on religious affiliation)`\
`[6]  Other Faiths`\
`[7]  Historically Black Protestant Churches`\
`[8]  Jehovah's Witness`\
`[9]  Catholic`\
`[10]  Buddhist`\
`[11]  Mormon`\
`[12]  Muslim`\
`[13]  Hindu`\
`[14]  Other Christian`\
`[15]  Orthodox`\
`[16]  Other World Religions`\

to:

`[1] Evangelical Prot        Mainline Prot           Unaffiliated`\
`[4] Jewish                  Don’t know/refused      Other Faiths`\
`[7] Historically Black Prot Jehovah's Witness       Atheist`\
`[10] Agnostic                Catholic                Buddhist`\
`[13] Mormon                  Muslim                  Hindu`\
`[16] Other Christian         Orthodox                Other World Religions`\

and replace income values:

`[1] 75 to under $100,000      20 to under $30,000       30 to under $40,000`\
`[4] Less than $10,000         50 to under $75,000       $150,000 or more`\
`[7] 40 to under $50,000       Don't know/Refused (VOL.) 100 to under $150,000`\
`[10] 10 to under $20,000`\

with:

`[1] "$75-100k"           "$20-30k"      "$30-40k"       "<$10k"`\
`[5] "$50-75k"            ">150k"        "$40-50k"       "Don't know/refused"`\
`[9] "$100-150k"          "$10-20k"`\

using the `stringr` library and dataframe column operations.

Once your strings are setup, factor the incomes, and use `count()` to create the tidy version.


```{=tex}
\begin{table}[H]
  \centering
  \input{data/pew-clean.tex}
  \caption{The first ten rows of the tidied Pew survey dataset on income and religion. The \texttt{column} has been renamed to \texttt{income}, and \texttt{value} to \texttt{freq}.}
  \label{tbl:pew-clean}
\end{table}
```
\newpage

### TB

For `tb`, starter code will take care of the first-step, minor, tidy-ups. Next, use `pivot_longer` to construct the molten version, before doing some additional clean-up/ rearrangement to reach the tidy format.

```{=tex}
\begin{table}[H]
  \centering
  \input{data/tb-raw.tex}
  \caption{Original TB dataset. Corresponding to each `m' column for males, there is also an `f' column for females, \texttt{f1524}, \texttt{f2534} and so on. These are not shown to conserve space. Note the mixture of 0s and missing values (---). This is due to the data collection process and the distinction is important for this dataset.}
  \label{tbl:tb-raw}
\end{table}
```
```{=tex}
\begin{table}[H]
  \centering
  \subfloat[Molten data]{\label{tbl:tb-molten}\input{data/tb-clean-1.tex}}%
  \hspace{2em}%
  \subfloat[Tidy data]{\label{tbl:tb-tidy}\input{data/tb-clean-2.tex}}
  
  \caption{Tidying the TB dataset requires first melting, and then splitting the \texttt{column} column into two variables: \texttt{sex} and \texttt{age}.}
  \label{tbl:tb-clean}
\end{table}
```
\newpage

### WEATHER

The weather data, being particularly fickle, uses a custom `read`  method provided in your starter code. Review the function and attempt to use the, base R, `read.fwf()` function to read in the raw data. What's the difference? Can `read.fwf()` be used instead?


```{=tex}
\begin{table}[H]
  \centering
  \input{data/weather-raw.tex}
  \caption{Original weather dataset.  There is a column for each possible day in the month.  Columns \texttt{d9} to \texttt{d31} have been omitted to conserve space.}
  \label{tbl:weather-raw}
\end{table}
```

After reading in the raw data you'll need to do the following to create the above table and melt the data:

-   Subset for just the year 2010, and only elements `"TMAX"` and `"TMIN"` 
-   Remove all columns besides `id`, `year`, `month`, `element`, and columns beginning with `value` 
-   Remove the extra `0`s from the `id` column 
-   Change the column names beginning with `value`, ex `value_1`, to `d`, example `d1` 
-   Divide the values in the new `d` columns by 10 
-   Finally, make values of the `element` column lowercase

A quick `pivot_longer()` and voila, you've made `melt_weather`.

Translate the new `variable` column from `dn` to just `n`, and convert the year, month, and day data into a single column of dates.  Use `as.date()` and `ISOdate()` accordingly.

Don't forget to select and arrange your columns to match the table below.

One more thing to do now, use a tidyr pivot to construct the final `tidy_weather` dataframe!


```{=tex}
\begin{table}[H]
  \centering
  \subfloat[Molten data]{\label{tbl:weather-molten}\input{data/weather-clean-1.tex}}%
  \hspace{2em}%
  \subfloat[Tidy data]{\label{tbl:weather-tidy}\input{data/weather-clean-2.tex}}
  
  \caption{(a) Molten weather dataset. This is almost tidy, but instead of values, the \texttt{element} column contains names of variables. Missing values are dropped to conserve space. (b) Tidy weather dataset. Each row represents the meteorological measurements for a single day. There are two measured variables, minimum (\texttt{tmin}) and maximum (\texttt{tmax}) temperature; all other variables are fixed.}
  \label{tbl:weather-clean}
\end{table}
```

